---
title: "R - ggplot"
output: html_document
---

Diamonds data set and explore the relationship of diamond weight (size) with price. We would expect larger diamonds to cost more.

```{r}
# Load grid package
require("grid")

# Create a promise to bring Diamonds data into memory
data(diamonds)
```

Look at some of the probability distributions...
```{r}
hist_price <- ggplot(diamonds, aes(x=price))
hist_price <- hist_price + geom_histogram(
  aes(fill=factor(color), y=..density..), 
  binwidth = 250,
  position = "stack") + theme_gray(9) + 
  labs(x="Price", y="Density") +
  labs(title="Diamonds - Price") +
  theme(plot.margin = unit(rep(0,4), "lines")) +
  theme(legend.position = "none")
print(hist_price)

hist_carat <- ggplot(diamonds, aes(x=carat))
hist_carat <- hist_carat + 
  geom_histogram(aes(fill=factor(color), y=..density..), 
  binwidth = 0.1,
  position = "stack") + theme_gray(9) + 
  labs(x="Weight", y="Density") +
  labs(title="Diamonds - Carat") +
  theme(plot.margin = unit(rep(0,4), "lines")) +
  theme(legend.position = "none")
print(hist_carat)

color_prc <- ggplot(diamonds, aes(x=color))
color_prc + geom_bar(fill="gray", colour="gray") +
  labs(x="Color") + labs(title="Diamonds - Color")
```

```{r}

# Plot relationship of carat on price by color
carat_prc <- ggplot(diamonds, aes(x=carat,y=price))
carat_prc + labs(title="Diamonds - Weight to Price by Color") +
  theme(plot.title = element_text(size = rel(2), colour = "blue")) +
  labs(x="Weight", y="Price") + 
  geom_point(aes(color = factor(color)))
```

The relationship is not linear. If we are considering a linear regression model to predict price then we should see if we can transform the variables and create a more linear relationship.

```{r}
# Remove the non-linearity of the relationship
diamonds$logprice <- log(diamonds$price)
diamonds$logcarat <- log(diamonds$carat)
diamonds$logdepth <- log(diamonds$depth)
diamonds$logx <- log(diamonds$x+1)
diamonds$logy <- log(diamonds$y+1)
diamonds$logz <- log(diamonds$z+1)

cor(diamonds$logcarat, diamonds$logprice)
cor(diamonds$logdepth, diamonds$logprice)
cor(diamonds$table, diamonds$logprice)
cor(diamonds$logx, diamonds$logprice)
cor(diamonds$logy, diamonds$logprice)
cor(diamonds$logz, diamonds$logprice)

diamonds$vol <- diamonds$logx*diamonds$logy*diamonds$logz
cor(diamonds$vol, diamonds$logprice)

cor(diamonds$vol, diamonds$carat)

lcarat_prc <- ggplot(diamonds, aes(x=logcarat,y=logprice))
lcarat_prc + labs(title="Diamonds - Weight to Price (Linear)") +
  theme(plot.title = element_text(size = rel(2), colour = "blue")) +
  labs(x="Weight", y="Price") + 
  geom_point(aes(color = factor(color)))

ldepth_prc <- ggplot(diamonds, aes(x=logdepth,y=logprice))
ldepth_prc + labs(title="Diamonds - Depth to Price (Linear)") +
  theme(plot.title = element_text(size = rel(2), colour = "blue")) +
  labs(x="Depth", y="Price") + 
  geom_point(aes(color = factor(color)))
```

Now we will fit a linear model to see if logcarat can be used to predict logprice.

```{r}
# Fit a linear model
fit <- lm(logprice ~ logcarat, data=diamonds)
str(fit)
class(fit)
diamonds$presids <- resid(fit)
diamonds$pred <- predict(fit)

dtmodel1_prc <- ggplot(diamonds, aes(x=logcarat,y=logprice))
dtmodel1_prc <- dtmodel1_prc + 
  labs(title="Diamonds - Price and Weight with Predictions") +
  theme(plot.title = element_text(size = rel(2), colour = "blue")) +
  labs(x="Weight", y="Price") +
  geom_point(aes(color = factor(color))) +
  theme(legend.position="top") +
  geom_point(aes(color="red", y=pred))
print(dtmodel1_prc)


require(gvlma)
gvmodel <- gvlma(fit)
summary(gvmodel)
```

Let's plot the residuals and see if we can visualize the heteroscedasticity.

```{r}
dtcarat_prc <- ggplot(diamonds, aes(x=pred,y=presids))
dtcarat_prc <- dtcarat_prc + 
  labs(title="Diamonds - Residuals to Price and Weight") +
  theme(plot.title = element_text(size = rel(2), colour = "blue")) +
  labs(x="Predicted Price", y="Price Residuals") +
  geom_point(aes(color = factor(color))) +
  theme(legend.position="top")
print(dtcarat_prc)

errorhist <- ggplot(diamonds, aes(x=presids))
errorhist <- errorhist + 
  geom_histogram(aes(fill="gray", y=..density..), binwidth = 0.05) +
  theme_gray(9) + 
  labs(x="Error", y="Density") +
  labs(title="Diamonds Error Distribution") +
  theme(plot.margin = unit(rep(0,4), "lines")) +
  theme(legend.position = "none")
print(errorhist)
```

Assess the relationship between Color and Price
```{r}
# First, convert color to a numeric (integer) and test correlation
diamonds$ColorNum <- as.numeric(diamonds$color)
cor(diamonds$ColorNum, diamonds$price)

# Another way is to bin price and look at a cross tabulation with
# color
diamonds$BinPrice <- ifelse(diamonds$price < 950, 
        "1LOW", ifelse(diamonds$price <= 5324, "2MED", "3HIGH"))
table(diamonds$color, diamonds$BinPrice)
```

Assess the relationship between clarity and price
```{r}
diamonds$ClarityNum <- as.numeric(diamonds$clarity)
cor(diamonds$ClarityNum, diamonds$price)

table(diamonds$clarity, diamonds$BinPrice)
```

Let's try to refit the model, adding our volume variable
```{r}
fit2 <- lm(logprice ~ logcarat+vol, data=diamonds)
diamonds$presids2 <- resid(fit2)
diamonds$pred2 <- predict(fit2)

gvmodel2 <- gvlma(fit2)
summary(gvmodel2)

str(fit2)

fit$coefficients
fit2$coefficients
```

Let's look at the residuals
```{r}
dtcarat_prc2 <- ggplot(diamonds, aes(x=pred2,y=presids2))
dtcarat_prc2 <- dtcarat_prc + 
  labs(title="Model 2 - Residuals to Price by Weight and Volume") +
  theme(plot.title = element_text(size = rel(2), colour = "blue")) +
  labs(x="Price", y="Price Residuals") +
  geom_point(aes(color = factor(color))) +
  theme(legend.position="top")
print(dtcarat_prc2)
```

```{r}
# Create a density histogram of the price
# factor(color)
hist_price <- ggplot(diamonds, aes(x=price))
hist_price <- hist_price + geom_histogram(
  aes(colour=factor(color), y=..density..), 
  binwidth = 100,
  position = "stack") + theme_gray(9) + 
  labs(x=NULL, y=NULL) +
  theme(plot.margin = unit(rep(0,4), "lines")) +
  theme(legend.position = "none")

# Create a density histogram of the weight (carat)
hist_carat <- ggplot(diamonds, aes(x=carat))
hist_carat <- hist_carat + geom_histogram(
  aes(colour=factor(color), y=..density..), 
  binwidth = 0.05,
  position = "stack") + theme_gray(9) + 
  labs(x=NULL, y=NULL) +
  theme(plot.margin = unit(rep(0,4), "lines")) +
  theme(legend.position = "none")

# Use grid viewports to create a plot with overlaid subplots
subvp1 <- viewport(width=0.4, height=0.2, 
                   x=0.2775, y=0.20)
subvp2 <- viewport(width=0.4, height=0.2,
                   x=0.775, y=0.675)
dtcarat_prc
print(hist_price, vp=subvp1)
print(hist_carat, vp=subvp2)

# Use grid layout to create a scatterhist plot similar
# to MatLab
grid.newpage()
pushViewport(viewport(layout=grid.layout(10,10)))
vplayout <- function(x,y,a) 
  viewport(layout.pos.row = x,
           layout.pos.col = y,
           angle = a)
vp3=vplayout(1:8,3:10,0)
vp4=vplayout(7:8,3:7,90)
vp5=vplayout(9:10,4:9,0)
print(dtcarat_prc, vp=vp3)
print(hist_price, vp=vp4)
print(hist_carat, vp=vp5)
```


Following is an example of how to get a dataset from the web, load it into R and begin exploring the data.

```{r}
# UCI is a great resource for datasets. Here's an example of how to 
# retrieve the Abalone data
uciaddress <- "http://archive.ics.uci.edu/ml/machine-learning-databases/"
dataset <- "abalone/abalone.data"
getdataset <- paste(uciaddress, dataset, sep="")
abalone <- read.csv(getdataset)
str(abalone)

```

Now let's name the columns...

```{r}

colnames(abalone) <- c("Gender","Length","Diameter",
                       "Height", "Whole_wgt",
                       "Shucked_wgt", "Viscera wgt",
                       "Shell wgt", "Rings")
str(abalone)
```

Here's a method for get a random sample from a dataframe

```{r}
# Quick way to get a sample from a dataframe
randomSample = function(dataframe,n,replacement) {     
  return (dataframe[sample(nrow(dataframe), n, replace=replacement),]) }
  
absample<-randomSample(abalone, 40, FALSE)
```

Let's recode the gender attribute using our sample to demonstrate using nested ifelse(). We might want to do this to use Gender in a clustering algorithm.

```{r}
args(ifelse)
absample$GenderInt <- ifelse(absample$Gender == 'M', 1, 
       ifelse(absample$Gender == 'F', 2, 3))
absample$GenderInt
absample$Gender
```

Find the mean number of rings for each gender type
```{r}
?tapply
tapply(absample$Rings, absample$Gender, mean)

table(list(absample$Gender, absample$Rings))
```

How good is the sample relative to whole_wgt? How would you measure?
```{r}
summary(absample$Whole_wgt)
summary(abalone$Whole_wgt)
```

How is Shell Wgt referenced?
```{r}
summary(absample$Shell wgt)
summary(absample$"Shell wgt")
```

Create a new variable size as a binning of Whole_wgt
```{r}
absample$Size <- ifelse(absample$Whole_wgt < 0.85, 
                        "SMALL", "LARGE")
table(list(absample$Gender, absample$Size))
```

##Repeated measure data.
Here we have subject, age, height, and occasion
```{r}
# ggplot oxboys data
require(nlme, quiet = TRUE, warn.conflicts = FALSE)
data(Oxboys)
str(Oxboys)
```

Leave out the grouping variable
```{r}
p4 <- ggplot(Oxboys, aes(age, height))
p4 + geom_line()
```

Now properly group
```{r}
p4 <- ggplot(Oxboys, aes(age, height, group = Subject))
p4 + geom_line()
p4 + geom_smooth(aes(group = Subject), method = "lm", se = F)
```

Added a smooth line for each subject, wanted a smooth line for the entire data
```{r}
p4 <- ggplot(Oxboys, aes(age, height, group = Subject))
p4 <- p4 + geom_line()
p4 + geom_smooth(aes(group = 1), method = "lm", size = 2, se = F)

#Let's look at box plots
p5 <- ggplot(Oxboys, aes(Occasion, height))
p5 <- p5 + geom_boxplot()
print(p5)
p5 <- p5 + geom_line(aes(group = Subject), color="red")
print(p5)
```

##Plotting different data sets on different layers   

In the plots above, fitted lines for individuals doesn't use information about typical growth patterns and group fit ignores with-in subject correlation.   

We can try a mixed model using the nlme package   
```{r}
model <- lme(height ~ age, data = Oxboys, 
             random = ~ 1 + age | Subject)
str(model)

# Create base spaghetti plot
p6 <- ggplot(Oxboys, aes(age, height, group = Subject))
p6 <- p6 + geom_line()
print(p6)

# Get the predictions into a new data set
age_grid <- seq(-1, 1, length = 10)
subjects <- unique(Oxboys$Subject)
preds <- expand.grid(age = age_grid, Subject = subjects)
preds$height <- predict(model, preds)

# Add the prediction layer from the new data set on
# a new layer
p6 <- p6 + geom_line(data = preds, color = "blue", 
                     size = 0.4)
print(p6)
```

Now lets look at the residuals   

Add predictions to the original data
```{r}
Oxboys$fitted <- predict(model)
# Now calculate the residuals
Oxboys$resid <- with(Oxboys, fitted - height)
str(Oxboys)

# Now create the residuals plot
p7 <- ggplot(Oxboys, aes(age, resid, group = Subject))
p7 + geom_line() + geom_smooth(aes(group=1))
```

Looks like the residuals are not random indicating some signal that hasn't been captured in the model.  

Add a polynomial term and refresh the data...
```{r}
model2 <- update(model, height ~ age + I(age^2))
Oxboys$fitted2 <- predict(model2)
Oxboys$resid2 <- with(Oxboys, fitted2 - height)

# Now "refresh" the data in the base plot and re-plot
p7 %+% Oxboys + aes(y=resid2) + geom_line() + geom_smooth(aes(group=1))
```

