---
title: "R - ggplot"
output: html_document
---

#Plotting in R

R's base graphics: implemented in the same way as in the Bell Labs S3 system developed by Becker, Chambers, and Wilks.

__grid__ was designed by Paul Murrell to overcome some of these limitations and as a result packages like lattice, __ggplot2__, vcd or hexbin (on Bioconductor) use grid for the underlying primitives.

__Plotting:__ plotrix, vcd, hexbin (on Bioconductor ), gclus and gplots  
__Graphic Applications:__ This area aids in creating effective displays.  
•  Effect ordering: gclus,cba and seriation  
•  Large Data Sets: ash, hexbin, scagnostics  
•  Trees and Graphs: ape, ade4, igraph, diagram, Rgraphviz  
__Graphics Systems:__ lattice, __ggplot2__  
__Devices:__ cairoDevice, RGtk2, RSvgDevice, rgl, JavaGD.   
__Colors:__ colorspace, RColorBrewer. dichromat
__Interactive Graphics:__ rggobi (Ggobi), iplots, iwidgets, playwith  
__Development: grid, rgl, RoSuDA, gridBase  

CRAN Task View: Graphic Displays & Dynamic Graphics & Graphic Devices & Visualization

#ggplot2  
  
•  Based on the grammar of graphics (Wilkinson,2005)  
•  Hadley Wickham(2005+): Primacy of layers and adaptation for embedding in R  
•  Grammar tells us that a statistical graphic, drawn on a coordinate system, is a mapping from data to aesthetic attributes (color, shape, size) of geometric objects (points, lines, bars) potentially containing statistical transformations and facets of the data.  
   
Most graphic packages are a collection of special cases, __but ggplot is__   
•  Designed to be layered  
•  Produce graphics using structured thinking  
•  Encourages the creation of totally new graphics  
•  Easily combine existing plots  
•  Represent multiple data sets  

#ggplot2 Grammer  
•  data  
•  aesthetic mappings    
•  geoms  
•  stats  
•  scales  
•  coord system   

#What is a plot?  
1.  Default dataset and mappings from variables to aesthetics  
2.  One or more layers   
    • Geom   
    •	Stat  
    •	Position adjustment  
    •	Dataset  
    •	Aesthetic mappings (optional)  
3.  Scale  
4.  Coordinate system  
5.  Faceting specification  

#What is a layer
Responsible for creating objects perceived on a plot
•  Data and aesthetic mapping  
•  Statistical transformation (stat)   
•  Geometric object (geom)  
•  Position adjustment  

#What is a Scale
•  Controls mapping from data to aesthetic attributes  
•  Operates across all data in the plot  
•  Ensures consistent mapping  
•  Example: Color Gradient – Maps a segment of a real line to a path through the color space

#What is a Coordinate system
Maps the position of objects onto the plane of the plot.  
•  Cartesian most common  
•  Polar  
•  Various map projections  
•  Affect all position variables simultaneously  
•  Change the appearance of geometric objects  
•  Scaling is performed before statistical transformation, coordinate transformations after.  

#What is a Facet  
•  General case of conditioned or trellised plots  
•  Makes it easy to plot different subsets of the data  

#How to Render a ggplot  
•  Print()–Renders to the screen  
•  ggsave()–Renders to a disk  
•  save()–Saves a cached copy (including data!)  

#geoms
geom_abline geom_area geom_bar geom_bin2d geom_blank geom_boxplot geom_contour geom_crossbar geom_density geom_density2d geom_dotplot geom_errorbar geom_errorbarh
geom_freqpoly geom_hex geom_histogram geom_hline geom_jitter geom_line geom_linerange geom_map geom_path geom_point geom_pointrange geom_polygon geom_quantile
geom_raster geom_rect geom_ribbon geom_rug geom_segment geom_smooth geom_step geom_text geom_tile geom_violin geom_vline

#stats
stat_bin stat_bin2d stat_bindot stat_binhex stat_boxplot stat_contour stat_density stat_density2d stat_ecdf stat_function stat_identity stat_qq stat_quantile
stat_smooth stat_spoke stat_sum stat_summary stat_summary_hex stat_summary2d stat_unique stat_ydensity

#scales
expand_limits
guide_legend
guide_colourbar(guide_colorbar)
scale_alpha(scale_alpha_continuous, scale_alpha_discrete)
scale_area
scale_colour_brewer(scale_color_brewer, scale_fill_brewer)
scale_colour_gradient(scale_color_continuous, scale_color_gradient, scale_colour_continuous, scale_fill_continuous, scale_fill_gradient)
scale_colour_gradient2(scale_color_gradient2, scale_fill_gradient2) scale_colour_gradientn(scale_color_gradientn, scale_fill_gradientn) scale_colour_grey(scale_color_grey, scale_fill_grey)
scale_colour_hue(scale_color_discrete, scale_color_hue, scale_colour_discrete, scale_fill_discrete, scale_fill_hue)
scale_identity(scale_alpha_identity, scale_color_identity, scale_colour_identity, scale_fill_identity, scale_linetype_identity,
scale_shape_identity, scale_size_identity)
scale_manual(scale_alpha_manual, scale_color_manual, scale_colour_manual, scale_fill_manual, scale_linetype_manual, scale_shape_manual, scale_size_manual)
scale_linetype(scale_linetype_continuous, scale_linetype_discrete)
scale_shape(scale_shape_continuous, scale_shape_discrete)
scale_size(scale_size_continuous, scale_size_discrete)
scale_x_continuous(scale_x_log10, scale_x_reverse, scale_x_sqrt, scale_y_continuous, scale_y_log10, scale_y_reverse, scale_y_sqrt)
scale_x_date(scale_y_date) scale_x_datetime(scale_y_datetime) scale_x_discrete(scale_y_discrete) labs(ggtitle, xlab, ylab) update_labels
xlim(ylim)

#Creating ggplots

##ggplot __always__ operates (requires) a data frame

Let's look at a data frame
```{r}
require(ggplot2)
require(grid)
data(diamonds)
str(diamonds)
str(diamonds[1])
str(diamonds$carat)
str(diamonds[[1]])
str(diamonds$cut)
class(str(diamonds[2]))
class(str(diamonds[[2]]))
class(diamonds[[2]])
```

Let's create a simple bar graph of cut several ways
```{r}
cut_bargraph_1 <- ggplot(diamonds, aes(x=diamonds$cut))
str(cut_bargraph_1)
cut_bargraph_1 + geom_bar() 

cut_bargraph_2 <- ggplot(diamonds[2], aes(x=diamonds[[2]]))
str(cut_bargraph_2)
cut_bargraph_2 <- cut_bargraph_2 + geom_bar(fill="gray", colour="red") + labs(x=names(diamonds)[2])
str(cut_bargraph_2)
print(cut_bargraph_2)
str(names(diamonds))
class(names(diamonds)[2])
cut_bargraph_2[2]
str(colnames(diamonds)[2])
rownames(diamonds)
```

Now let's print a couple histograms (count)
```{r}
Test_Frame <- diamonds[1]
bin50 <- (max(Test_Frame[[1]])-min(Test_Frame[[1]]))/50
Test_Plot <- ggplot(Test_Frame, aes(x=Test_Frame[[1]]))
Test_Plot <- Test_Plot +
  geom_histogram(colour = 'black',
                 fill = 'gray',
                 binwidth=bin50) +
  labs(x=names(Test_Frame)[[1]])
print(Test_Plot)
```

Now modify Test_Plot to become a density histogram...
```{r}
Test_Plot <- Test_Plot + aes(y=..density..) 

Test_Plot + labs(y="Density")
```

Or just recreate Test_Plot...
```{r}
Test_Plot <- ggplot(Test_Frame, aes(x=Test_Frame[[1]]))
Test_Plot <- Test_Plot + aes(y=..density..) +
  geom_histogram(colour = 'black',
                 fill = 'gray',
                 binwidth=bin50) +
  labs(x=names(Test_Frame)[[1]])
print(Test_Plot)
```

Plotting variable bin sizes...
```{r}
Test_Frame <- diamonds
binsizes <- c(3, 10, 20, 50, 75, 100)
for(i in 1:length(binsizes)){
  bins <- (max(Test_Frame[[1]])-min(Test_Frame[[1]]))/binsizes[i]
  Test_Plot <- ggplot(Test_Frame, aes(x=Test_Frame[[1]]))
  Test_Plot <- Test_Plot + aes(y=..density..) +
    geom_histogram(colour = 'blue',
                   fill = 'gray',
                   binwidth=bins) +
    labs(x=names(Test_Frame)[[1]])
  print(Test_Plot)
}
```


Can we plot from within a function?
```{r}
Handle_Num <- function(Num_Frame) 
{
  bin50 <- (max(Num_Frame[[1]])-min(Num_Frame[[1]]))/50
  Num_Plot <- ggplot(Num_Frame, aes(x=Num_Frame[[1]]))
  Num_Plot <- Num_Plot + aes(y=..density..) +
    geom_histogram(colour = 'gray',
                   fill = 'gray',
                   binwidth=bin50) +
    labs(x=names(Num_Frame)[[1]])
  print(Num_Plot)
}


Handle_Num(diamonds[1])

#Hmmm weird error - looks like my indexing is right.
#Let's google ggplot2 Error in eval(expr, envir, enclos) : object 'Num_Frame' not found 

#Added work around for scoping bug in ggplot
Handle_Num2 <- function(Num_Frame) 
{
  bin50 <- (max(Num_Frame[[1]])-min(Num_Frame[[1]]))/50
  Num_Plot <- ggplot(Num_Frame, aes(x=Num_Frame[[1]]),
                    environment = environment())
  Num_Plot <- Num_Plot + aes(y=..density..) +
    geom_histogram(colour = 'gray',
                   fill = 'gray',
                   binwidth=bin50) +
    labs(x=names(Num_Frame)[[1]])
  print(Num_Plot)
}

Handle_Num2(diamonds[1])

#Add loop in function for all variables, bin sizes and mean line
Plot_Num <- function(Num_Frame, Bin_Sizes) 
{
  for (i in 1:length(Num_Frame))
  {
    for (bins in 1:length(Bin_Sizes))
    {
      #Calculate bin size
      num_bins <- (max(Num_Frame[[i]])-min(Num_Frame[[i]]))/Bin_Sizes[bins]
      
      #Calculate the mean and label
      col_mean <- mean(Num_Frame[[i]])
      mean_Label = sprintf("%8.2f ", col_mean)
      
      #Plot regular count histogram
      Num_Plot <- ggplot(Num_Frame, aes(x=Num_Frame[[i]]),
                         environment = environment())
      Num_Plot <- Num_Plot +
        geom_histogram(colour = 'gray',
                       fill = 'gray',
                       binwidth=num_bins) +
        labs(x=names(Num_Frame)[[i]]) +
        geom_vline(xintercept=col_mean, colour='red') +
        annotate("text",x=col_mean,y=0,label=mean_Label,hjust=0)
      print(Num_Plot)
      
      #Plot denisty histogram
      print(Num_Plot + aes(y=..density..) + labs(y="Density"))
#       Num_Plot <- ggplot(Num_Frame, aes(x=Num_Frame[[i]]),
#                          environment = environment())
#       Num_Plot <- Num_Plot + aes(y=..density..) +
#         geom_histogram(colour = 'gray',
#                        fill = 'gray',
#                        binwidth=num_bins) +
#         labs(x=names(Num_Frame)[[i]]) +
#         geom_vline(xintercept=col_mean, colour='red') +
#         annotate("text",x=col_mean,y=0,label=mean_Label,hjust=0)
#       print(Num_Plot)
    } 
  }
}

#Test it
Plot_Num(diamonds[1], c(5,50))
```




